Estimating dark matter halo concentration using the integrated mass profile

### 2015-11-06 paper/paper.tex
descripcion metodo

\begin{equation}
\rho(r) = \frac{\rho_c\delta_c}{r/r_s(1+r/r_s)^2},
\label{eq:definition}
\end{equation}
%
where $\rho_c\equiv 3H^2/8\pi G$ is the Universe critical density, $H$
is the Hubble constant, $G$ and the universal gravitational constant,
$\delta_c$ is the halo dimensionless characteristic density and $r_s$
is the scale radius. This radius marks {\bf the point where the
  logarithmic slope of the density profile is equal to -2} marking the
transition between the power law scaling $\rho\propto r^{-1}$ for
$r<r_s$ and $\rho\propto r^{-3}$ for $r>r_s$.

@@

In this paper we use $\Delta_h=740$ roughly corresponding to 200 times
the critical density {\bf at redshift z=0}.

@@

From this normalization value and for later convenience we define the
following function
%
\begin{equation}
f(x) = \ln\left(1+x\right)-\left(\frac{x}{x+1}\right).
\label{eq:f_NFW}
\end{equation}
%
The most interesting feature of Eq. (\ref{eq:profile}) is that the
concentration is the only free parameter to describe the integrated
mass profile.

@@

\begin{equation}
v(<x)=\sqrt{\frac{1}{A}\left[\frac{\ln\left(1+xc\right)}{x}-\frac{c}{xc+1}\right]},
\end{equation}

This normalized profile always shows a maximum provided that the
concentration is larger than $c>2$.
It is possible to show that for the NFW profile the maximum is
provided by

\begin{equation}
\mathrm{max}(v(<x)) = \sqrt{\frac{c}{x_{\rm max}}\frac{f(x_{\rm
      max})}{f(c)}},
\label{eq:max_v}
\end{equation}
where $x_{\rm max}=2.163$ \citep{Klypin2014} and the function $f(x)$
was defined in Eq. (\ref{eq:f_NFW}).

@@

\subsection{Estimates from the density and velocity profiles}
{\bf To date,} there are two standard methods to estimate
concentrations in dark matter halos extracted from N-body simulations.

The first method tries to directly estimate the density profile.  It
takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  Then, it estimates the
density in each logarithmic bin counting the particles and dividing by
the corresponding shell volume.  At this point is is possible to make
a direct fit to the density as a function of the radial coordinate.
This method has been {\bf used broadly for more than two decades} to
study the mass-concentration-redshift relation of dark matter halos
using the Millennium Simulation Series.

{\it no veo la necesidad de citar el ``trabajo mas'' reciente que usa
  el metodo, cuando ha sido usado durante mas de dos decadas!}

@@

\subsection{Our proposal: estimate from the integrated mass profile}
Our method uses the integrated mass profile.  First we define the
center of the halo to be at the position of the particle with the
lowest gravitational potential.  Then we rank the particles by their
increasing radial distance from the center.  From this ranked list of
$i=1,N$ particles, the total mass at a radius $r_i$ is $M_i=i\times
m_p$, where $r_i$ is the position of the $i$-th particle and $m_p$ is
the mass of a single computational particle. In this process we
discard the particle at the center.

We divide the enclosed mass mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter. We use an Affine Invariant
Markov chain Monte Carlo implemented in the python module {\em emcee}
\citep{emcee} to sample the likelihood function distribution defined
by ${\cal L}(c)\propto \exp(-\chi^2(c)/2)$ where the $\chi^2(c)$ is
written as

@@

The construction of the numerical integrated mass profile has two
advantages with respect to the density and velocity methods, it does
not involve any binning (unlike the density methods ) and the sampling
method uses the information from all the particles in the halo (unlike
the velocity method).

@@

We generate in total $400$ mock halos split into four different groups
of $100$ halos each.  The four groups differ in the total number of
particles for their halos: $20$, $200$, $2000$ and $20000$.  Inside
each group the halos have random concentration values in the range
$1<c<20$ with a uniform distribution.  For all these halos {\bf ww?}
the concentration values using the density, velocity and mass methods
described in the previous section.  We quantify the difference between
the expected $c_{in}$ and obtained $c_{out}$ values by
%
\begin{equation}
D=(c_{in}-c_{out})/c_{in},
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, and
$\left|{\cal{H}}\right|$ is the number of haloes in ${\cal{H}}$. {\bf
  The larger the value of $\avg{|D|}$, the larger the discrepace of
  the concentrations obtained for that method}

>>>>

\begin{equation}
\rho(r) = \frac{\rho_c\delta_c}{r/r_s(1+r/r_s)^2},
\label{eq:definition}
\end{equation}
%
where $\rho_c\equiv 3H^2/8\pi G$ is the Universe critical density, $H$
is the Hubble constant, $G$ and the universal gravitational constant,
$\delta_c$ is the halo dimensionless characteristic density and $r_s$
is the scale radius. 
This radius marks the point where the logarithmic slope of the density
profile is equal to -2, marking the transition between the power law
scaling $\rho\propto r^{-1}$ for $r<r_s$ and $\rho\propto r^{-3}$ for
$r>r_s$.

@@

In this paper we use $\Delta_h=740$ roughly corresponding to 200 times
the critical density at redshift z=0.

@@

From this normalization and for later convenience we define the
following function
%
\begin{equation}
f(x) = \ln\left(1+x\right)-\left(\frac{x}{x+1}\right).
\label{eq:f_NFW}
\end{equation}
%
The most interesting feature of Eq. (\ref{eq:profile}) is that the
concentration is the only free parameter to describe the integrated
mass profile.

@@

\begin{equation}
v(<x)=\sqrt{\frac{1}{A}\left[\frac{\ln\left(1+xc\right)}{x}-\frac{c}{xc+1}\right]},
\end{equation}
%
This normalized profile always shows a maximum provided that the
concentration is larger than $c>2$.
It is possible to show that for the NFW profile the maximum is
provided by

\begin{equation}
\mathrm{max}(v(<x)) = \sqrt{\frac{c}{x_{\rm max}}\frac{f(x_{\rm
      max})}{f(c)}},
\label{eq:max_v}
\end{equation}
where $x_{\rm max}=2.163$ \citep{Klypin2014} and the function $f(x)$
correspondes to the definition in Eq. (\ref{eq:f_NFW}).

@@

\subsection{Estimates from the density and velocity profiles}

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.

The first method tries to directly estimate the density profile.  
It takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.  
At this point is possible to make a direct fit to the density as a function of the radial coordinate.
This method has been used broadly for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

@@

\subsection{Our proposal: estimate from the integrated mass profile}

Our method uses the integrated mass profile.  
First we define the center of the halo to be at the position of the
particle with the lowest gravitational potential.  
Then we rank the particles by their increasing radial distance from
the center. 
From this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is the position of the $i$-th
particle and $m_p$ is the mass of a single computational particle. In
this process we discard the particle at the center.

We divide the enclosed mass mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter. 
We use an Affine Invariant Markov Chain Monte Carlo implemented in the
python module {\em emcee} \citep{emcee} to sample the likelihood
function distribution defined by ${\cal L}(c)\propto
\exp(-\chi^2(c)/2)$ where the $\chi^2(c)$ is 
written as

@@

The construction of the numerical integrated mass profile has two
advantages with respect to the density and velocity methods, it does
not involve any binning (unlike the density methods) and the sampling
method uses the information from all the particles in the halo (unlike
the velocity method).

@@

We generate in total $400$ mock halos split into four different groups
of $100$ halos each.  The four groups differ in the total number of
particles for their halos: $20$, $200$, $2000$ and $20000$.  
Inside each group the halos have random concentration values in the range
$1<c<20$ with a uniform distribution.  
For all these halos we find  the concentration values using the
density, velocity and integrated mass methods described in the
previous section.

We quantify the difference between the expected $c_{in}$ and measured
$c_{out}$ values by 
%
\begin{equation}
D=(c_{in}-c_{out})/c_{in},
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, and
$\left|{\cal{H}}\right|$ is the number of haloes in ${\cal{H}}$. 
  The larger the value of $\avg{|D|}$, the larger the discrepace of
  the concentrations obtained for that method.
